---
title: "Notas Prueba 2 (SOL114)"
author: "Mauricio Bucca"
format: html
editor: visual
---

## Notas

```{r, echo=F, message=FALSE, warning=FALSE}
library("pacman")
p_load(googlesheets4, tidyverse, kableExtra,mclust)

#gs4_auth()

sheet_url_1 <- "https://docs.google.com/spreadsheets/d/1BFWZx0b44gvvfTgA337I30oJwiV09mkF3_bUTM-WepE/edit#gid=1064220691"

sheet_url_2 <- "https://docs.google.com/spreadsheets/d/1BFWZx0b44gvvfTgA337I30oJwiV09mkF3_bUTM-WepE/edit#gid=1548544822"

notas_prueba1 <- gs4_get(sheet_url_1)
notas_prueba2 <- gs4_get(sheet_url_2)

notas_prueba1<- read_sheet(notas_prueba1, sheet= "Prueba 1")



notas <- read_sheet(notas_prueba2, sheet= "Prueba 2")

data_grades <- notas %>% drop_na(Nota, Rut) %>% filter(Nota > 0) %>% 
  select(Rut,`1.i`:`Nota_curved`) %>% 

  mutate(`P(Nota < tunota)` =rank(Nota_curved)/n()) 

data_grades %>%
  mutate(Nota = Nota_curved) %>% select(-Nota_curved) %>%
  knitr::kable(digits=2) %>% kable_styling(font_size = 9)

```

## Resumen estadístico

El siguiente gráfico muestra la distribución de las notas:

```{r, echo=F, message=FALSE, warning=FALSE}

# Compute statistical metrics

media <- mean(data_grades$Nota_curved)
mediana <- median(data_grades$Nota_curved)
sd <- sd(data_grades$Nota_curved)
min_val <- min(data_grades$Nota_curved)
max_val <- max(data_grades$Nota_curved)

# Set limits for y-axis so the annotations have enough space
limits_y <- c(0, max(density(data_grades$Nota_curved)$y)*1.5)

# Define ggplot
p <- ggplot(data_grades, aes(x = Nota_curved)) +
  geom_histogram(aes(y = ..density..), 
                 fill = "blue", 
                 alpha = 0.5, 
                 bins = 30) +
  geom_vline(xintercept = media, linetype = "dashed", color = "black") +
  geom_text(aes(x = 1.1, y = limits_y[2] * 0.9, 
                label = paste("Media =", round(media, 2))), hjust = 0) +
  geom_text(aes(x = 1.1, y = limits_y[2] * 0.8, 
                label = paste("Min =", round(min_val, 2))), hjust = 0) +
  geom_text(aes(x = 1.1, y = limits_y[2] * 0.7, 
                label = paste("Max =", round(max_val, 2))), hjust = 0) +
  geom_text(aes(x = 1.1, y = limits_y[2] * 0.6, 
                label = paste("SD =", round(sd, 2))), hjust = 0) +
  labs(title = "Distribución de Notas**",
       x = "Nota",
       y = "Densidad") +
  xlim(1,8) +
  theme_minimal() +
  coord_cartesian(ylim = limits_y)

# Display plot
print(p)
```

Las notas fueron re-escaladas para final el 7 en la nota máxima obtenida.
Formalmente:

$$
 \text{Nota Final} =  \text{Nota Original} \times \frac{\text{7}}{\max(\text{Nota Original})} =\text{Nota Original} \times \frac{\text{7}}{6.6}
$$

Usando un Gaussian mixture model encontramos 3 clusters de notas:

```{r, echo=F, message=FALSE, warning=FALSE}

# Set seed for reproducibility
set.seed(7615)

# Fit a Gaussian mixture model with the "EM" algorithm and adjusted bandwidth
fit <- Mclust(data_grades$Nota_curved, G = 3, adjustment = list(diagCov = TRUE))

# Get the cluster assignments for each data point
cluster_assignments <- fit$classification

# Calculate the means for each cluster
cluster_means <- tapply(data_grades$Nota_curved, cluster_assignments, mean)

# Create a ggplot without a legend, add horizontal lines, annotate cluster means, and add vertical lines
p <- ggplot(data_grades, aes(x = Nota, fill = factor(cluster_assignments))) +
  geom_density(alpha = 0.5) +
  labs(title = "Gaussian Mixture Model",
       x = "Nota",
       y = "Density") +
  theme_minimal() +
  scale_fill_manual(name = "Component", 
                    values = c("#E41A1C",  # Red
                                "#377EB8",  # Blue
                                "#4DAF4A",  # Green
                                "#984EA3")) +  # Purple
  theme(legend.position = "none") +  # Remove the legend
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +  # Add horizontal line at y = 0
  geom_text(data = data.frame(cluster = unique(cluster_assignments), 
                             cluster_mean = cluster_means),
            aes(x = cluster_mean, y = 2, label = round(cluster_mean, 1), fill = factor(cluster)),
            vjust = -0.5, size = 3) +  # Annotate cluster means with colors and rounded values
  geom_vline(data = data.frame(cluster_mean = cluster_means),
             aes(xintercept = cluster_mean), color = "black", linetype = "dotted")  # Add vertical lines

# Display the plot
print(p)


```

Desagregando por pregunta (sin ajustar):

```{r, echo=F, message=FALSE, warning=FALSE}

# Loading necessary libraries
library(ggridges)

# Data preparation
data_long <- data_grades %>%  select(Rut,`1.i`:`8`) %>%
  pivot_longer(cols = `1.i`:`8`, names_to = "Variable", values_to = "Value")

# Calculate means for annotations
data_long %>%
  group_by(Variable) %>%
  summarise(Promedio = mean(Value, na.rm = TRUE),
            Min = min(Value, na.rm = TRUE),
            Max = max(Value, na.rm = TRUE),
            SD = sd(Value, na.rm = TRUE)
            ) %>% 
  knitr::kable(digits=2) %>% kable_styling(font_size = 9)
  

```

```{r, echo=F, message=FALSE, warning=FALSE}

# Assuming you have your datasets notas_prueba1 and data_grades
# Joining datasets and renaming columns
data_combined <- data_grades %>%
  rename(Nota_2 = Nota_curved) %>%
  select(Rut, Nota_2) %>%
  left_join(notas_prueba1 %>% select(Rut, Nota), by="Rut") %>%
  rename(Nota_1 = Nota) 
```

Comparando el rendimiendo en las Prueba 1 y Prueba 2, observamos que un `r 100* round(mean( data_combined$Nota_2 > data_combined$Nota_1, na.rm=T),2)`% de los estudiantes mejoró su nota.
Adicionalmente, existe una asociación positiva pero relativamente débil entre ambas.

```{r, echo=F, message=FALSE, warning=FALSE}
# Plotting
plot <- ggplot(data_combined, aes(x = Nota_1, y = Nota_2)) +
  geom_point(aes(colour = Nota_2 > Nota_1)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") + # 45-degree line
  geom_smooth(method="lm", se = FALSE, color = "black") + # Linear regression line
  labs(x = "Nota Prueba 1", y = "Nota Prueba 2", color = "Nota 2 > Nota 1") +
  theme_minimal() +
  scale_color_manual(values = c("#E41A1C", "#377EB8"))

# Calculate correlation coefficient
cor_coef <- cor(data_combined$Nota_1, data_combined$Nota_2, use = "pairwise.complete.obs")

# Add annotation for correlation coefficient
plot <- plot + annotate("text", 
                        x = 2.3,
                        y = 6.7, 
                        label = paste("corr =", round(cor_coef, 2)), hjust = 0, vjust = 1)

# Print the plot
print(plot)

```
