---
title: "Probabilidad e Inferencia Estadística"
subtitle: "#9: Teoremas Asintóticos"
author: "<br> Mauricio Bucca <br> [github.com/mebucca](https://github.com/mebucca) <br> mebucca@uc.cl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["gentle-r.css","xaringan-themer.css"]
    df_print: default
    nature:
      ratio: 16:9
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: true
      slideNumberFormat: "%current%"
editor_options: 
  
  chunz_output_type: console
---  
class: inverse, center, middle

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(tidyverse)
library(xaringanthemer)
style_duo_accent(primary_color ="#4B6F44", secondary_color ="#F4C430",
                 background_color = "#f8f7f3",
                 header_font_google = google_font("Archivo"),
                 text_font_google   = google_font("Inconsolata"), 
                 link_color= "#C05A50"

)
```



#Teoremas Asintóticos


---
class: inverse, center, middle

#Ley de los Grandes Números
## Convergencia en Probabilidad


---
## Ley de los Grandes Números



- $X$ es una variable aleatoria con primer momento $\mathbb{E}(X) = \mu$

- Generamos/tomamos una muestra $\{X_{1},X_{2}, \dots, X_{n}\}$ de tamaño $n$ donde:
  
    - Cada $X_{i}$ sigue la misma distribución que la variable aleatoria $X$
    - Las $X_{i}$'s son independientes entre si

--

- La media aritmética en .bold[esta muestra] es  $\bar{X}_{n} = \frac{1}{n}\sum_{i=1}^{n}X_i$

- $\bar{X}_{n}$ es .bold[aleatoria] porque las $X_{i}$'s son aleatorias.


<br>
--

.bold[Ley de los Grandes Números] establece que, a medida que el tamaño de una muestra aleatoria crece, la media aritmética de la muestra .bold[converge en probabilidad] al promedio teórico/poblacional: $\bar{X} \overset{p}{\to} \mathbb{E}(X)$.



.content-box-primary[
.bolder[Matemáticamente,]
$$\color{white}{\lim_{n \to \infty}\mathbb{P}\left( \left| \bar{X} - \mu \right| \geq \epsilon \right) = 0}$$
]


---
## Ley de los Grandes Números

.bold[Convergencia en Probabilidad]
.pull-left[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=7}
# Load ggplot2 paczage for visualization
library(ggplot2)

# Initialize a vector to store the sample means
sample_means <- c()

# Initialize an empty dataframe for the plot
plot_data <- data.frame(x = integer(0), y = double(0))

# epsilon value for the lines
epsilon <- 0.15

# Plot configuration with ggplot2
p <- ggplot() +
  geom_line(data = plot_data, aes(x = x, y = y), color = "#4B6F44") +
  xlab("n: tamaño muestra (# dados lanzados)") +
  ylab(expression(paste(bar(X)[n], ": promedio muestral"))) + # Updated y-axis label
  ylim(2.5, 4.5) + 
  xlim(1, 6000) + 
  geom_segment(aes(x = 1, y = 3.5, xend = 5000, yend = 3.5), color = "#4B6F44") + 
  geom_segment(aes(x = 1, y = 3.5 + epsilon, xend = 5000, yend = 3.5 + epsilon), color = "#C05A50", linetype = "dashed") +
  geom_segment(aes(x = 1, y = 3.5 - epsilon, xend = 5000, yend = 3.5 - epsilon), color = "#C05A50", linetype = "dashed") +
  annotate("text", x = 5550, y = 3.5, label = "µ = 3.5", size = 5, color = "#4B6F44") +  # Added text for µ
  annotate("text", x = 3000, y = 3.5 + epsilon + 0.15, label = expression(paste("µ + ", epsilon)), size = 5, color = "#C05A50") +  # Added text for upper epsilon line
  annotate("text", x = 3000, y = 3.5 - epsilon - 0.15, label = expression(paste("µ - ", epsilon)), size = 5, color = "#C05A50") +  # Added text for lower epsilon line
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 22),
    axis.text.x = element_text(size = 22),
    axis.title.y = element_text(size = 24),
    axis.title.x = element_text(size = 24),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  )

# Loop to simulate the Law of Large Numbers with dice rolling
for (n in 1:5000) {
  # Roll a die n times
  sample <- sample(1:6, size = n, replace = TRUE)
  
  # Calculate the sample mean
  sample_mean <- mean(sample)
  
  # Add the sample mean to the vector of sample means
  sample_means <- c(sample_means, sample_mean)
  
  # Add data to the dataframe for the plot
  plot_data <- rbind(plot_data, data.frame(x = n, y = sample_mean))
  
  # Update the plot
  if (n %% 5000 == 0) {
    print(p + geom_point(data = plot_data, aes(x = x, y = y), size = 0.5, color = "#4B6F44", alpha=0.15) +
            geom_line(data = plot_data, aes(x = x, y = y), size = 0.5, color = "#4B6F44", alpha=0.15)
    )
  }
  
  # Pause for 0.01 seconds between each iteration
  Sys.sleep(0.001)
}

```
]


.pull-right[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=7}
# Load ggplot2 package for visualization
library(ggplot2)

# Initialize a dataframe for the plot
plot_data <- data.frame(sample_size = integer(0), mean_outside_epsilon = double(0))

# ε (epsilon) value for the range
epsilon <- 0.15

# True mean of rolling a 6-sided die
true_mean <- 3.5

# Sample sizes to consider (skipping 0 as it's not a meaningful sample size for this context)
sample_sizes <- seq(1, 6000, by = 50)

# Loop through different sample sizes
for (sample_size in sample_sizes) {
  
  # Counter for means outside the ε range
  count_outside_epsilon <- 0
  
  # Draw 1000 samples for each sample size
  for (i in 1:100) {
    
    # Roll a die "sample_size" times
    sample <- sample(1:6, size = sample_size, replace = TRUE)
    
    # Calculate the sample mean
    sample_mean <- mean(sample)
    
    # Check if the sample mean is outside the ε range
    if (abs(sample_mean - true_mean) > epsilon) {
      count_outside_epsilon <- count_outside_epsilon + 1
    }
  }
  
  # Calculate percentage of sample means outside ε range
  percentage_outside_epsilon <- (count_outside_epsilon / 1000) * 100
  
  # Add data to the dataframe for the plot
  plot_data <- rbind(plot_data, data.frame(sample_size = sample_size, mean_outside_epsilon = percentage_outside_epsilon))
  
}

# Plotting
ggplot(data = plot_data, aes(x = sample_size, y = mean_outside_epsilon)) +
  geom_line(color = "#4B6F44", size = 0.8, alpha=0.7) +
  geom_point(color = "#4B6F44", size = 1.2) +
  xlab("n: tamaño muestra (# dados lanzados)") +
  ylab(expression(paste("Porcentaje de ", bar(X)[n], "'s en (", -epsilon, ", ", epsilon, ")" ))) +
  theme_minimal() +
  xlim(1,6000) + 
  theme(
    axis.text.y = element_text(size = 22),
    axis.text.x = element_text(size = 22),
    axis.title.y = element_text(size = 24),
    axis.title.x = element_text(size = 24),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  )

```
]

---
class: inverse, center, middle

#Valor Esperado y Varianza
#de la
#Media Muestral

---
##Valor Esperado de la Media Muestral

.pull-left[

- $X$ es una variable aleatoria con primer momento $\mathbb{E}(X) = \mu$

- Generamos/tomamos una muestra $n$ del tipo:


- $\{X_{1},X_{2}, \dots, X_{n}\}$

- Las $X_{i}$'s son .bold[iid]
    
- En cada muestra calculamos la .bold[media muestral],  $\bar{X}_{n}$

- $\bar{X}_{n}$ es .bold[aleatoria] porque las $X_{i}$'s son aleatorias.

<br>

.bold[¿Cuál es el promedio de dichas medias muestrales?]

]

--

.pull-right[

\begin{align}
\mathbb{E}(\bar{X}_{n}) &= \mathbb{E}\big(\frac{1}{n} \sum^{n}_{i=1} X_{i} \big) \\ \\
&=\frac{1}{n}\mathbb{E}\big( X_{1} + X_{2} + \cdots + X_{n} \big) \\ \\
&=\frac{1}{n}\big( \mathbb{E}(X_{1}) + \mathbb{E}(X_{2}) + \cdots  + \mathbb{E}(X_{n})\big) \\ \\
&=\frac{1}{n}\big( \mu + \mu + \cdots + \mu \big) \\ \\
&=\frac{1}{n}\big( n \mu \big) \\ \\
&=\mu
\end{align}
]

---
##Valor Esperado de la Media Muestral

Para cualquier $n$ el promedio de esas las diferentes $\bar{X}_{n}$'s es igual a $\mathbb{E}(X)$.

.center[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=7}

# Load ggplot2 package for visualization
library(ggplot2)

# Initialize a dataframe for the individual sample means and another for the averages
sample_means_data <- data.frame(sample_size = integer(0), sample_mean = double(0))
averages_data <- data.frame(sample_size = integer(0), avg_sample_mean = double(0))

# True mean of rolling a 6-sided die
true_mean <- 3.5

# Sample sizes to consider
sample_sizes <- seq(1, 6000, by = 50)

# Loop through different sample sizes
for (sample_size in sample_sizes) {
  
  sum_sample_means <- 0
  
  # Draw 1000 samples for each sample size
  for (i in 1:100) {
    
    # Roll a die "sample_size" times
    sample <- sample(1:6, size = sample_size, replace = TRUE)
    
    # Calculate the sample mean
    sample_mean <- mean(sample)
    sum_sample_means <- sum_sample_means + sample_mean
    
    # Add individual sample mean to the dataframe
    sample_means_data <- rbind(sample_means_data, data.frame(sample_size = sample_size, sample_mean = sample_mean))
  }
  
  # Calculate average of sample means for the current sample size
  avg_of_sample_means <- sum_sample_means / 100
  
  # Add average to the averages dataframe
  averages_data <- rbind(averages_data, data.frame(sample_size = sample_size, avg_sample_mean = avg_of_sample_means))
}

# Plotting
ggplot() +
  geom_line(data = averages_data, aes(x = sample_size, y = 3.5, color = "a"), size = 1.5) +   
  geom_line(data = averages_data, aes(x = sample_size, y = avg_sample_mean, color = "b"), size = 1.5, alpha=0.5) +
  scale_color_manual(
    values = c("a" = "#4B6F44", "b" = "#C05A50"), 
    labels = c(expression(E(bar(X)[n]) == mu), expression(paste("Promedio de ",bar(X)[n],"'s"))
  )) +
  xlab("n: tamaño muestra (# dados lanzados)") +
  ylab(expression(bar(X)[n])) + 
  ylim(2.5, 4.5) + 
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 22),
    axis.text.x = element_text(size = 22),
    axis.title.y = element_text(size = 24),
    axis.title.x = element_text(size = 24),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = c(0.95, 0.95),
    legend.justification = c("right", "top"),
    legend.background = element_blank(),
    legend.box.background = element_rect(color="transparent", fill = "transparent"),
    legend.box.margin = margin(0,0,0,0),
    legend.title = element_blank(),
    legend.key = element_blank() # Removes the border around the legend items
  )

```
]


---
## Varianza y Desviación Standard de la Media Muestral

.pull-left[

- $X$ es una variable aleatoria con primer momento $\mathbb{E}(X) = \mu$

- Generamos/tomamos una muestra $n$ del tipo:


- $\{X_{1},X_{2}, \dots, X_{n}\}$

- Las $X_{i}$'s son .bold[iid]
    
- En cada muestra calculamos la .bold[media muestral],  $\bar{X}_{n}$

- $\bar{X}_{n}$ es .bold[aleatoria] porque las $X_{i}$'s son aleatorias.

<br>

.bold[¿Cuanto varían dichas medias muestrales?]

]

--

.pull-right[

\begin{align}
\text{Var}(\bar{X}) &= \text{Var}\big(\frac{1}{n} \sum^{n}_{i=1} X_{i} \big) \\
&= \frac{1}{n^2} \text{Var}\big( \sum^{n}_{i=1} X_{i} \big) \\ \\
&=\frac{1}{n^2}\big( \text{Var}(X_{1}) + \text{Var}(X_{2}) + \cdots + \text{Var}(X_{n})\big) \\ \\
&=\frac{1}{n^2}\big( \sigma^2 + \sigma^2 + \cdots + \sigma^2 \big) \\ \\
&=\frac{1}{n^2}\big( n \sigma^2 \big) \\ \\
&=\frac{\sigma^2}{n} \implies \quad \sqrt{\text{Var}(\bar{X})}  = \frac{\sigma}{\sqrt{n}}
\end{align}
]



---
## Varianza de la Media Muestral

A medida que aumenta $n$ los diferentes $\bar{X}_{n}$'s se concentran más cerca del promedio de las $\bar{X}_{n}$'s (que, a su vez, coincide con $\mathbb{E}(X)$).

.center[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=7}

# Load ggplot2 package for visualization
library(ggplot2)

# Initialize a dataframe for the individual sample means and another for the averages
sample_means_data <- data.frame(sample_size = integer(0), sample_mean = double(0))
averages_data <- data.frame(sample_size = integer(0), avg_sample_mean = double(0))

# True mean of rolling a 6-sided die
true_mean <- 3.5

# Sample sizes to consider
sample_sizes <- seq(1, 6000, by = 50)

# Loop through different sample sizes
for (sample_size in sample_sizes) {
  
  sum_sample_means <- 0
  
  # Draw 1000 samples for each sample size
  for (i in 1:100) {
    
    # Roll a die "sample_size" times
    sample <- sample(1:6, size = sample_size, replace = TRUE)
    
    # Calculate the sample mean
    sample_mean <- mean(sample)
    sum_sample_means <- sum_sample_means + sample_mean
    
    # Add individual sample mean to the dataframe
    sample_means_data <- rbind(sample_means_data, data.frame(sample_size = sample_size, sample_mean = sample_mean))
  }
  
  # Calculate average of sample means for the current sample size
  avg_of_sample_means <- sum_sample_means / 100
  
  # Add average to the averages dataframe
  averages_data <- rbind(averages_data, data.frame(sample_size = sample_size, avg_sample_mean = avg_of_sample_means))
}

# Plotting
ggplot() +
  geom_point(data = sample_means_data, aes(x = sample_size, y = sample_mean, color = "Individual Sample Mean"), size = 0.5, alpha=0.05) +
  geom_line(data = averages_data, aes(x = sample_size, y = 3.5, color = "Average of Sample Means"), size = 1) +
  scale_color_manual(
    values = c("Individual Sample Mean" = "#4B6F44", "Average of Sample Means" = "#C05A50"), 
    labels = c(expression(E(bar(X)[n]) == mu), expression(bar(X)[n]))
  ) +
  xlab("n: tamaño muestra (# dados lanzados)") +
  ylab(expression(bar(X)[n])) + 
  ylim(2.5, 4.5) + 
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 22),
    axis.text.x = element_text(size = 22),
    axis.title.y = element_text(size = 24),
    axis.title.x = element_text(size = 24),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = c(0.95, 0.95),
    legend.justification = c("right", "top"),
    legend.background = element_blank(),
    legend.box.background = element_rect(color="transparent", fill = "transparent"),
    legend.box.margin = margin(0,0,0,0),
    legend.title = element_blank(),
    legend.key = element_blank() # Removes the border around the legend items
  )
```
]


---
## Varianza de la Media Muestral

A medida que aumenta $n$ la desviación standard de $\bar{X}_{n}$'s .bold[disminuye a una taza decreciente] específica.

.center[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=7}

library(ggplot2)

# True variance of rolling a 6-sided die
variance <- 2.9166667

# Data frame for the standard deviations
sd_data <- data.frame(sample_size = integer(0), sample_sd = double(0), theoretical_sd = double(0))

# Sample sizes to consider
sample_sizes <- seq(1, 600, by = 5)

# Loop through different sample sizes
for (sample_size in sample_sizes) {
  
  sample_means <- numeric(100) # array to store the sample means for each sample
  
  # Draw 100 samples for each sample size
  for (i in 1:100) {
    
    # Roll a die "sample_size" times
    sample <- sample(1:6, size = sample_size, replace = TRUE)
    
    # Calculate the sample mean and store it
    sample_means[i] <- mean(sample)
  }
  
  # Calculate the standard deviation of the sample means
  sd_sample_means <- sd(sample_means)
  
  # Compute theoretical SD
  theoretical_sd <- sqrt(variance)/sqrt(sample_size)
  
  # Add to the data frame
  sd_data <- rbind(sd_data, data.frame(sample_size = sample_size, sample_sd = sd_sample_means, theoretical_sd = theoretical_sd))
}

# Plotting
ggplot(data = sd_data, aes(x = sample_size)) +
  geom_line(aes(y = sample_sd, color = "sample_sd"), size = 1) +
  geom_line(aes(y = theoretical_sd, color = "theoretical_sd"), linetype = "dashed", size = 1) +
  scale_color_manual(
    values = c("sample_sd" = "#4B6F44", "theoretical_sd" = "#C05A50"),
    labels = c(expression(Desviación~estándar~de~bar(X)[n]), expression(sqrt(Var(bar(X)[n])) == sigma/sqrt(n))
    )) +
  xlab("n: tamaño muestra (# dados lanzados)") +
  ylab(expression(Desviación~estándar~de~bar(X)[n])) + 
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 14),
    axis.title.y = element_text(size = 16),
    axis.title.x = element_text(size = 16),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.title = element_blank(),
    legend.position = c(0.95, 0.95),
    legend.justification = c("right", "top"),
    legend.background = element_blank()
  ) +
  annotate("text", x = 500, y = 0.25, label = expression(paste(frac("d", dn) * (sigma/sqrt(n)), " = ", frac(-sigma, 2 * sqrt(n)^3))), parse = TRUE)

```
]

---
## Corolario

Típicamente contamos con .bold[sólo una muestra], con un $n$ fijo. 

.bold[¿"Cuán bien" nuestro estimado muestral] $\bar{X}_{n}$ .bold[captura el parámetro verdadero E(X)?] 

--
.center[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=6, fig.width=7}

library(ggplot2)

# Initialize a dataframe for the individual sample means and another for the averages
sample_means_data <- data.frame(sample_size = integer(0), sample_mean = double(0))
averages_data <- data.frame(sample_size = integer(0), avg_sample_mean = double(0))

# True mean of rolling a 6-sided die
true_mean <- 3.5

# Sample sizes to consider
sample_sizes <- seq(1, 6000, by = 50)

# Loop through different sample sizes
for (sample_size in sample_sizes) {
  
  sum_sample_means <- 0
  
  # Draw 1000 samples for each sample size
  n_sims=100
  for (i in 1:n_sims) {
    
    # Roll a die "sample_size" times
    sample <- sample(1:6, size = sample_size, replace = TRUE)
    
    # Calculate the sample mean
    sample_mean <- mean(sample)
    sum_sample_means <- sum_sample_means + sample_mean
    
    # Add individual sample mean to the dataframe
    sample_means_data <- rbind(sample_means_data, data.frame(sample_size = sample_size, sample_mean = sample_mean))
  }
  
  # Calculate average of sample means for the current sample size
  avg_of_sample_means <- sum_sample_means / n_sims
  
  # Add average to the averages dataframe
  averages_data <- rbind(averages_data, data.frame(sample_size = sample_size, avg_sample_mean = avg_of_sample_means))
}

# Select one data point that is farther from the true mean
far_from_mean <- sample_means_data[abs(sample_means_data$sample_mean - true_mean) > 0.4, ]
selected_far <- sample(1:nrow(far_from_mean), 1)

# Select 4 more random points excluding the one chosen above
remaining_rows <- setdiff(1:nrow(sample_means_data), as.numeric(rownames(far_from_mean)[selected_far]))
selected_remaining <- sample(remaining_rows, 4)

selected_rows <- c(as.numeric(rownames(far_from_mean)[selected_far]), selected_remaining)

# Plotting
ggplot() +
  geom_point(data = sample_means_data, aes(x = sample_size, y = sample_mean, color = "Individual Sample Mean"), size = 0.5, alpha=0.05) +
  geom_line(data = averages_data, aes(x = sample_size, y = 3.5, color = "Average of Sample Means"), size = 1) +
  geom_point(data = sample_means_data[selected_rows,], aes(x = sample_size, y = sample_mean), color = "black", size = 3) + 
  scale_color_manual(
    values = c("Individual Sample Mean" = "#4B6F44", "Average of Sample Means" = "#C05A50"), 
    labels = c(expression(E(bar(X)[n]) == mu), expression(bar(X)[n]))
  ) +
  xlab("n: tamaño muestra (# dados lanzados)") +
  ylab(expression(bar(X)[n])) + 
  ylim(2.5, 4.5) + 
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 22),
    axis.text.x = element_text(size = 22),
    axis.title.y = element_text(size = 24),
    axis.title.x = element_text(size = 24),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = c(0.95, 0.95),
    legend.justification = c("right", "top"),
    legend.background = element_blank(),
    legend.box.background = element_rect(color="transparent", fill = "transparent"),
    legend.box.margin = margin(0,0,0,0),
    legend.title = element_blank(),
    legend.key = element_blank() # Removes the border around the legend items
  )

``` 
]

---
class: inverse, center, middle


##Hasta la próxima clase. Gracias!


<br>
Mauricio Bucca <br>
https://mebucca.github.io/ <br>
github.com/mebucca




